Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-11-22 manuscript.qmd
add missing citations

This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [cite 2024 graph extension paper]. Below, we briefly recap the unilateral model to introduce the notation required for extending the framework to a bilateral model, described in @sec-ext-to-contra. For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021; cite 2024 graph extension paper].

@@

The transition matrix $\mathbf{A}$ is parameterized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network. Edges from the primary tumor to an LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Arcs from a LNL $v$ to a LNL $r$ represent the probability $t_{vr}$ for inter-nodal progression. In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].

@@

Within the cohort of patients with oropharyngeal tumors, subgroups can be identified whose tumors are located in different subsites, e.g. at the base of the tongue or on the tonsils. The lymphatic metastatic spread patterns of these subgroups may differ slightly. If so, the model could benefit from more fine-grained information about the precise tumor subsite for its training and prediction. We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [cite ICCR paper, @zora231470]. The additionally represents an approach to extend the model towards tumor locations in the oral cavity, hypopharynx, and larynx.

>>>>

This paper builds on the previously developed unilateral model for ipsilateral lymph node involvement presented in [@ludwig_modelling_2024]. Below, we briefly recap the unilateral model to introduce the notation required for extending the framework to a bilateral model, described in @sec-ext-to-contra. For further details on the ipsilateral model, refer to earlier publications [@ludwig_hidden_2021; @ludwig_modelling_2024].

@@

The transition matrix $\mathbf{A}$ is parameterized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network. Edges from the primary tumor to an LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Arcs from a LNL $v$ to a LNL $r$ represent the probability $t_{vr}$ for inter-nodal progression. In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [@ludwig_modelling_2024].

@@

Within the cohort of patients with oropharyngeal tumors, subgroups can be identified whose tumors are located in different subsites, e.g. at the base of the tongue or on the tonsils. The lymphatic metastatic spread patterns of these subgroups may differ slightly. If so, the model could benefit from more fine-grained information about the precise tumor subsite for its training and prediction. We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [@noauthor_proceedings_2024, @zora231470]. The additionally represents an approach to extend the model towards tumor locations in the oral cavity, hypopharynx, and larynx.
